---
title: "lm and glm and la course"
author: "Steven Wink"
date: "8 november 2016"
output: html_document
---
course part of master "Mathematics: Statistical Science for the Life and Behavioural Sciences, 2016-2017"  

### linear models, general linear models and linear algebra  
## This course is part of the master provided by the mathematics institute of the university of Leiden    

## course material  
* Fox, J (2008), Applied Regression Analysis and Generalized Linear Models, Sage  
* Faraway, J.J. (2002), Practical Regression and Anova using R - web text: https://cran.r-project.org/doc/contrib/Faraway-PRA.pdf   
* Faraway, J.J. (2005) Extending the Linear Model with R, Chapman & Hall/ CRC
* Optional: McCulloch, C.E., Searle, S.R., and Neuhaus, J.M. (2008), Generalized, Linear, and Mixed Models, Wiley  

Note to self: There are newer versions of these books, check if OK to use the newer versions  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercises week 1 day 1  

```{r install load and explore packages faraway and car }
install.packages(c("faraway", "car"))
require(faraway)
require(car)

ls("package:faraway")
ls("package:car")
?qqPlot
help(package = "faraway") # functions and datasets for Books by Julian Faraway
help(package = "car") # companion to applied regression: Functions and Datasets to Accompany J. Fox and S. Weisberg, An R Companion to Applied Regression, Second Edition, Sage, 2011.

```


```{r 2. Follow analysis pages 10-13}
data(pima)
head(pima)
summary(pima)
sort(pima$diastolic)

 pima$diastolic[pima$diastolic == 0] <- NA
 pima$glucose[pima$glucose == 0] <- NA
 pima$triceps[pima$triceps == 0] <- NA
 pima$insulin[pima$insulin == 0] <- NA
 pima$bmi[pima$bmi == 0] <- NA

pima$test <- factor(pima$test)
summary(pima) 
hist(pima$diastolic)
plot(density(pima$diastolic,na.rm=TRUE))
plot(sort(pima$diastolic),pch=".")
plot(diabetes ~ diastolic, pima)
plot(diabetes ~ test,pima)

pairs(pima, col = pima$test)

```

```{r Exercise 3 faraway book }

# a numeric and graphical summary of prostate dataset
head(prostate)
summary(prostate)
pairs(prostate)

```


```{r Exercise D2.1 (fox) }
# a)
data(Sahlins)
head(Sahlins)
plot( Sahlins$acres ~Sahlins$consumers)

# weak positive linear relation
# two outliers, at ~(1.2, 3.1) and at ~(1.6, 1.3)

# b)
Sahlins$group <- c(rep("g1", 7), rep("g2", 6), rep("g3", 7))
groupMeans <-lapply( 
    split(Sahlins, Sahlins$group), 
    function(x) colMeans(x[, 1:2])) 
groupMeans <- t( as.data.frame(groupMeans) )
points(groupMeans, col = c("blue", "green", "red"))
legend( 2.1,3, legend = rownames(groupMeans), fill =  c("blue", "green", "red"), pch = ".")

# connect the group means with a simple nonparametric regression line:
lines(groupMeans)
# does not help to interpret relation between these two variables, no.

# c)
# remove outliers from g1 and g2:

Sahlins <- Sahlins[-4, ]
rownames(Sahlins) <- 1: nrow(Sahlins)
Sahlins <- Sahlins[-11, ]
rownames(Sahlins) <- 1: nrow(Sahlins)


groupMeans <-lapply( 
    split(Sahlins, Sahlins$group), 
    function(x) colMeans(x[, 1:2])) 
groupMeans <- t( as.data.frame(groupMeans) )
plot( Sahlins$acres ~Sahlins$consumers)
points(groupMeans, col = c("blue", "green", "red"))

legend( 2.1,3, legend = rownames(groupMeans), fill =  c("blue", "green", "red"), pch = ".")
lines(groupMeans)
# shifted away from outliers, relation between variables now more apparent

# d) locally weighted regression
lines( lowess( Sahlins$acres ~ Sahlins$consumers), col = "pink") 

```

```{r D2.2 Robey}
# a)
data(Robey)
head(Robey)
plot( Robey$tfr ~ Robey$contraceptors)

# strong negative linear relation
# Not really

# b)
Robey <- Robey[ order(Robey$contraceptors), ]
Robey$group <- c(rep("g1", 17), rep("g2", 16), rep("g3", 17))
groupMeans <-lapply( 
    split(Robey, Robey$group), 
    function(x) colMeans(x[, 3:2])) 

groupMeans <- t(as.data.frame(groupMeans) )

points(groupMeans, col = c("blue", "green", "red"))

legend( 60,7, legend = rownames(groupMeans), fill =  c("blue", "green", "red"), pch = ".")

# connect the group means with a simple nonparametric regression line:
lines(groupMeans)
# does not help to interpret relation between these two variables, no.

# c)
# remove outliers from g1 and g2:


# d) locally weighted regression
lines( lowess( Robey$tfr ~ Robey$contraceptors), col = "pink") 

```

chapter 3
```{r D3.1 }
# distributions. Symmetry, skewness, non-normality/ apparent normality, number of modes and presence or absence of unusual values
# histogram, qq plots boxplots scatterplots, pairs
head(Freedman)
hist(Freedman$population) # right tailed, no normality
hist(Freedman$nonwhite)# right tailed, no normality
hist(Freedman$density)# right tailed, no normality
hist(Freedman$crime) # normally distr, symmetrical

qqPlot(Freedman$population)# heavy right tailed. light left tailed
qqPlot(Freedman$nonwhite)#heavy right tailed. light left tailed
qqPlot(Freedman$density)# heavy right tailed. light left tailed
qqPlot(Freedman$crime) # normal

boxplot(Freedman) # same
plot(Freedman)
pairs(Freedman)

```


```{r exerc 7 }
rsample <- rnorm(20, mean = 170, sd = 10)
qqPlot(rsample)
rsample <- rnorm(200, mean = 170, sd = 10)
qqPlot(rsample)

```


```{r D3.4}
head(Burt)
iqLow <- lm(IQbio ~IQfoster, data = Burt[Burt$class =="low",])
iqMedium <- lm(IQbio ~IQfoster, data = Burt[Burt$class =="medium",])
iqHigh <- lm(IQbio ~IQfoster, data = Burt[Burt$class =="high",])
plot(Burt$IQfoster, Burt$IQbio)
abline(iqLow, col = "red")
abline(iqMedium, col = "green")
abline(iqHigh, col = "blue")

# yes, the slopes are too similar for a random sample dataset
```

Chapter 4

```{r D4.1 }
head(Freedman)
#population: Total 1968 population, 1000s.
#nonwhite: Percent nonwhite population, 1960.
#density: Population per square mile, 1968.
#crime: Crime rate per 100,000, 1969.

# nonwhite and crime are proportions

plot(density(Freedman$population, na.rm = TRUE))
qqPlot(Freedman$population)

# mainly shrink the right tail and expand the close to 0 values.
plot(density(log(Freedman$population), na.rm = TRUE))
qqPlot(log(Freedman$population))
# log is still too skewed. Try a different power rule
# Box-Cox transformation: with p 0 < 1
p <- -0.8
plot(density( 
  ((Freedman$population^p -1) / p), na.rm = TRUE))

qqPlot(((Freedman$population^p -1) / p))

p <- -0.8
plot(density( 
  Freedman$population^p , na.rm = TRUE))
qqPlot(Freedman$population^p)

# non white


plot(density(Freedman$nonwhite, na.rm = TRUE))
qqPlot(Freedman$nonwhite)

plot(density(log(Freedman$nonwhite), na.rm = TRUE))
qqPlot(log(Freedman$nonwhite))# log is fine

# now consider a logit, since it is a proportional value

plot(density(
      log(Freedman$nonwhite/100) / ( 1- (Freedman$nonwhite)/100)
  , na.rm = TRUE))

qqPlot( log(Freedman$nonwhite/100) / ( 1- (Freedman$nonwhite)/100))# log is fine
# also works

# density

plot(density(Freedman$density, na.rm = TRUE))
qqPlot(Freedman$density)

plot(density(log(Freedman$density), na.rm = TRUE))
qqPlot(log(Freedman$density))
# log fine

# crime
plot(density(Freedman$crime, na.rm = TRUE)) # already normal
qqPlot(Freedman$crime)

# already good

# consider using logit transformation (0-1 --> -inf - inf)

plot(density(
      log(Freedman$crime/100000) / ( 1- (Freedman$crime)/100000)
  , na.rm = TRUE))

qqPlot( log(Freedman$crime/100000) / ( 1- (Freedman$crime)/100000))# log is fine

# logit made it worse


```


```{r UN}
head(UN)
help(UN)
boxplot(UN$infant.mortality) # fraction infant deaths (per 1000)

qim <- quantile(UN$infant.mortality, na.rm=TRUE)

ratioFun <- function(x) { 
qim <- quantile(x, na.rm = TRUE)
  result <-  (qim[4] - qim[3]) / (qim[3]-qim[2])
  names(result) <- NULL
  return(result)
  }

ratioFun(UN$infant.mortality) # 2

# logit
p <- UN$infant.mortality /1000
ratioFun(log( p / (1-p)) ) # 0.88
boxplot(log( p / (1-p)))
plot(density(log( p / (1-p)), na.rm=TRUE))
plot(density(p, na.rm=TRUE))

# Box-Cox transformation
UN$infant.mortality

boxCoxFun <- function(x, p) {
transfx <-  (x^p - 1 / p)
transfx
}

for(i in 1:10){
  p <- seq(from = -1, to =1 , length.out=10)[i]
ratioOut <- ratioFun(
  boxCoxFun(UN$infant.mortality, p)

  )
print(paste0("p = ", p))
print(ratioOut)
}

boxplot(boxCoxFun(UN$infant.mortality, 0.11))
plot(density(boxCoxFun(UN$infant.mortality, 0.11), na.rm=TRUE))



```



```{r 11 Prestige }

head(Prestige)
plot(Prestige$prestige ~ Prestige$income)
lines(lowess(Prestige$prestige ~ Prestige$income))

# 4rth quadrant: downsize y, increase x
plot(Prestige$prestige ~ sqrt(Prestige$income))
lines(lowess(Prestige$prestige ~ sqrt(Prestige$income)))
lm(Prestige$prestige ~ sqrt(Prestige$income))
abline(a = 2.17, b = 0.564, lty = 2)

``` 

```{r 12 }
head(UN)
plot(UN$infant.mortality ~ UN$gdp)
plot(log(UN$infant.mortality) ~ log(UN$gdp))
abline(lm(log(UN$infant.mortality) ~ log(UN$gdp)), lty = 2)
UN <- na.omit(UN)
lines(lowess(log(UN$infant.mortality) ~ log(UN$gdp)))

``` 


```{r 13}

x <- seq(0.01, 0.99, length.out=100)
ylogit <- log( x / (1-x) )
plot(x,ylogit) # -inf and + inf near 0 and 1, respectively 

yprobit <- qnorm(x)

points(x, yprobit, col  ="red")

yarcsinsqrt <- (asin(sqrt(x)))

points(x,yarcsinsqrt, col = "green")
```


## Exercises week 1 day 2  
Exercises chapter 5
```{r 1 }

# a) by hand: B = 3/2, A = 1
# b)
XY <- data.frame(X=c(1, 2, 3), Y = c(2, 5, 5))
# A = Ybar - B*Xbar
# B = sum( (XY$X - mean(XY$X)) * (XY$Y - mean(XY$Y)) )
B <- sum( (XY$X - mean(XY$X)) * (XY$Y - mean(XY$Y)) ) / sum( (XY$X - mean(XY$X))^2 )
A <- mean(XY$Y) - B * mean(XY$X)

# c)
lm(Y~X, data = XY)

```


```{r D5.1}
#a) 
Sahlinsub <- Sahlins[ c(1,5,10,15,20),]
plot(Sahlinsub$consumers, Sahlinsub$acres)
#b)
SahRegr <- lm( Sahlinsub$acres ~ Sahlinsub$consumers)
abline( lm( Sahlinsub$acres ~ Sahlinsub$consumers) )
# A is the intercept, when zero consumers are in a household the acres of land of the household estimation
# Per unit of consumer per gardener, an increase of 0.4 acres per gardener exists based on the model of this subset of the data.

#c)
n = nrow(Sahlinsub)
k = 1

Yhat <- predict(SahRegr)
A <- coefficients(SahRegr)[1] 
B <- coefficients(SahRegr)[2] 
A + B* Sahlinsub$consumers
RSS <- sum((Sahlinsub$acres - Yhat)^2)
(SE <- sqrt( RSS / ( n - (k+1) )))
RSS <- sum((Sahlinsub$acres - Yhat)^2)

 TSS <- sum((Sahlinsub$acres - mean( Sahlinsub$acres ))^2) # from the null model with only intercept
 RSS
 
RegSS <- TSS - RSS
RSq <- RegSS / TSS 
r <- sqrt(RSq) # slope of regression is positive
 # the Standard error of the regression is the goodness of fit of the regression to the data. 

``` 


```{r D5.3}
# find A, B, SE and r
?Freedman
head(Freedman)
# response variable: crime
# explanatory variables: population & nonwhite & density
plot( crime ~ log(population), data = Freedman)
lmo <- lm( crime ~ log(population), data = Freedman  )
abline(lmo) 
coefficients(lmo)
summary(lmo)



```
